### 2022-04-23 misc/expo/tensorrank.tex
Added Matej

\def\FF{{\bf F}}
\def\T{{\rm T}}
\def\bias{\op{\rm bias}}
\def\prk{\op{\rm prk}}
\def\ark{\op{\rm ark}}
\def\bar{\overline}
\def\hat{\widehat}
\def\norm#1{|\!|#1|\!|}
\def\bignorm#1{\big|\!\big|#1\big|\!\big|}
\def\Norm#1{\Big|\!\Big|#1\Big|\!\Big|}
\def\normm#1{\bigg|\!\bigg|#1\bigg|\!\bigg|}
\def\x{{\bf x}}
\def\y{{\bf y}}

@@

\maketitle{The analytic rank of a tensor}{by}{Marcel K. Goh}{13 April 2022}

\floattext4.5 \ninebf Note. \ninepoint These notes are more or less a retelling of
some results in a 2019 paper entitled ``The analytic rank of tensors and its applications'' by
S.~Lovett. I go more deeply into definitions (for my own sake) so for other students this may be easier to
follow than the original paper.

@@

\proclaim Lemma \advthm. Let $d\ge 1$ and for each $I\subseteq [d]$, let $R_I : V^I\to \FF$
be an order-$|I|$ tensor. Consider the function
$$R(\x) = \sum_{I\subseteq [d]} R_I(x_I).$$
Then
$$\bigl|\bias(R)\bigr| \le \bias(R_{[d]}).$$

@@

\proclaim Theorem \advthm. Let $T,S: V^d\to \FF$ be order-$d$ tensors. Then
$$\ark(T+S) \le \ark(T) + \ark(S).$$

>>>>

\def\FF{{\bf F}}
\def\T{{\rm T}}
\def\bias{\op{\rm bias}}
\def\prk{\op{\rm prk}}
\def\ark{\op{\rm ark}}
\def\bar{\overline}
\def\hat{\widehat}
\def\norm#1{|\!|#1|\!|}
\def\bignorm#1{\big|\!\big|#1\big|\!\big|}
\def\Norm#1{\Big|\!\Big|#1\Big|\!\Big|}
\def\normm#1{\bigg|\!\bigg|#1\bigg|\!\bigg|}
\def\x{{\bf x}}
\def\y{{\bf y}}
\def\b{{\bf b}}

@@

\maketitle{The analytic rank of a tensor}{by}{Marcel K. Goh}{22 April 2022}

\floattext4.5 \ninebf Note. \ninepoint These notes are more or less a retelling of
some results in a 2019 paper entitled ``The analytic rank of tensors and its applications'' by
S.~Lovett. I rearranged results in an order that makes more sense to me.
I also go more deeply into definitions (for my own sake) and skip fewer steps,
so for other students this may be easier to
follow than the original paper.

@@

\edef\normbound{\the\thmcount}
\proclaim Lemma \advthm. Let $d\ge 1$ and for each $I\subseteq [d]$, let $R_I : V^I\to \FF$
be an order-$|I|$ tensor. Consider the function
$$R(\x) = \sum_{I\subseteq [d]} R_I(x_I).$$
Then
$$\bigl|\bias(R)\bigr| \le \bias(R_{[d]}).$$

@@

Before proving subadditivity, we introduce another bit of notation. For $I\subseteq [d]$,
let
$$T_I(\x,\y) = T(\x_I, y_{I^c}) = T(z_1, \ldots, z_d),$$
where $z_i = x_i$ if $i\in I$ and $z_i = y_i$ if $i\notin I$. After expanding out the definition
of multilinearity, we see that $T(\x+\y)$ decomposes as
$$ T(\x+\y) = \sum_{I\subseteq[d]} T_I(\x,\y).$$

\edef\subadditivity{\the\thmcount}
\proclaim Theorem \advthm. Let $T,S: V^d\to \FF$ be order-$d$ tensors. Then
$$\ark(T+S) \le \ark(T) + \ark(S).$$

\proof It is enough to show that
$$\bias(T+S) \ge \bias(T)\bias(S).$$
We express
$$\eqalign{
\bias(T)\bias(S) &= \Bigl(\ex_{\x\in V^d} \chi\bigl(T(\x)\bigr)\Bigr)
                    \Bigl(\ex_{\y\in V^d} \chi\bigl(T(\y)\bigr)\Bigr)\cr
&= \ex_{\x\in X} \ex_{\y\in Y} \chi\bigl( T(\x) + S(\y)\bigr) \cr
&= \ex_{\x\in X} \ex_{\y\in Y} \chi\bigl( T(\x) + S(\x+\y)\bigr) \cr
&= \bias\bigl( T(\x) + S(\x+\y)\bigr), \cr
}$$
and by our earlier decomposition the tensor product of a sum, this gives us
$$\bias(T)\bias(S)
= \bias \Bigl( T(\x) + \sum_{I\subseteq [d]} S_I(\x,\y) \Bigr)
= \bias \Bigl( (T+S)(\x) + \sum_{I\ne [d]} S_I(\x,\y) \Bigr), $$
Setting $\y = \b\in V^d$ to maximise the norm of this right-hand side, we have
$$\bias(T)\bias(S) \le \Bigl| \bias\Bigl( (T+S)(\x) + \sum_{I\ne [d]} S_I(\x, \b)\Bigr)\Bigr|.$$
Here $S_I(\x,\b)$ is an order-$|I|$ tensor in the variables $\x_I$. Thus we may apply
the previous lemma with
$$R_{[d]}(\x) = (T+S)(\x)\qquad\hbox{and}\qquad R_I(\x_I) = S_I(\x,\b)\ \hbox{for}\ I\ne [d]$$
to get $\bias(T)\bias(S) \le \bias(T+S)$.\slug

As an application of this theorem, we show that common roots of tensors on a common input are
positively correlated.

\proclaim Corollary \advthm. Let $T_1,\ldots,T_m, S_1,\ldots,S_n : V^d\to \FF$ be order-$d$ tensors.
Then
$$\eqalign{
\pr_{x\in V^d}&\bigl\{ T_1(x) = \cdots = T_m(x) = S_1(x) = \cdots = S_n(x) = 0\bigr\}\ge \cr
&\pr_{x\in V^d}\bigl\{ T_1(x) = \cdots = T_m(x) = 0\}\cdot
\pr_{x\in V^d}\bigl\{ S_1(x) = \cdots = S_n(x) = 0\}
}$$

\proof Define order-$(d+1)$ tensors $T$ and $S$ by setting
$$T(x_0,x_1,\ldots,x_d) = \sum_{i=1}^m x_{0,i} T_i(x_1,\ldots, x_d)$$
and
$$S(x_0,x_1,\ldots,x_d) = \sum_{i=m+1}^{m+n} x_{0,i} S_{i-m}(x_1,\ldots,x_d).$$
The left-hand side of the desired inequality is $\bias(T+S)$ and the right-hand side is
$\bias(T)\bias(S)$.\slug

\advsect Analytic rank and partition rank

In this section, we shall discuss properties of the analytic rank that allow it to be used as
the partition rank was used above in the cap-set problem. First, we show that the analytic rank
is bounded above by the partition rank.

\proclaim Theorem \advthm. Let $T:V^d\to\FF$ be an order-$d$ tensor. Then $\ark(T)\le \prk(T)$.

\proof By Theorem~\subadditivity, we only need to consider the case where $T$ has partition rank one.
So we assume that $T:V^d\to \FF$ factors as
$$T(\x) = T_1(\x_A)T_2(\x_B),$$
where $A\cup B = [d]$ is a nontrivial partition of $[d]$. We shall show that $\ark(T)\le 1 = \prk(T)$ by
showing that $\bias(T) \ge 1/|\FF|$. For $a,b\in \FF$ define the function
$$ F_{a,b}(\x) = \bigl(T_1(\x_A) + a\bigr) \bigl( T_2(\x_B) + b\bigr).$$
Letting $R_{[d]} = T$, $R_{\emptyset} = ab$, $R_A = bT_1(\x_A)$, $R_B = aT_2(\x_B)$, and $R_I$ be the
zero tensor (of the corresponding order)
for all other $I\subseteq [d]$, we see that $R = F_{a,b}$ and
by Lemma~\normbound, $\bigl|\bias(F_{a,b})\bigr| \le \bias(T)$.
On the other hand, we have
$$\eqalign{
\ex_{a,b\in \FF} \bias(F_{a,b}) &= \ex_{a,b\in \FF} \ex_{\x\in V^d}
   \chi\bigl( (T_1(\x_A) + a)(T_2(\x_B) + b) \bigr) \cr
&= \ex_{a,b\in F} \chi(ab) \cr
&= \pr_{b\in \FF} \{b=0\} \cr
&= {1\over |\FF|}. \cr
}$$
So we may find $a,b$ such that $\bias(T) \ge \bigl| \bias(F_{a,b})\bigr| \ge 1/|\FF|$, which
is what we wanted.\slug

Since we are often interested in finding some upper bound on the partition rank in applications,
this theorem suggests that it should actually be easier to upper bound the analytic
rank, though {\it ad hoc} methods to do so have not currently been developed.

In applications, it is also important that the partition rank of a tensor $T$
is bounded below by the size of an
independent set in $T$. Towards showing a version of this for the analytic rank, we first show
that the analytic rank does not increase when the tensor is restricted to a subspace.

\proclaim Lemma \advthm. Let $T : V^d \to \FF$ be an order-$d$ tensor and let $U\subseteq V$ be a
subspace. If $T|_U : U^d\to \FF$ is the restriction of $T$ to $U$, then $\ark(T|_U) \le \ark(T)$.

\proof Let $W\subseteq V$ be a subspace so that $V = U\oplus W$. Then any $v\in V$ has a unique
expression as $v = u+w$ for $u\in U$ and $w\in W$, so
$$\bias(T) = \ex_{u_1,\ldots,u_d\in U} \ex_{w_1,\ldots,w_d\in W}\chi\bigl(T(u_1+w_1,\ldots,u_d+w_d)\bigr).$$
Fix an arbitrary choice of $w_1,\ldots,w_d\in W$. Then
$$T(u_1+w_1,\ldots, u_d+w_d) = \sum_{I\subseteq [d]} T_I(u_I, w_{I^c}),$$
where $T_I(u_I, w_{I^c}) = T(z_1,\ldots,z_d)$ where $z_i = u_i$ if $i\in I$ and $z_i = w_i$ if $i\notin I$.
By Lemma~\normbound, we have
$$\eqalign{
\bigl| \ex_{u_1,\ldots,u_d\in U} \chi\bigl( T(u_1+w_1,\ldots,u_d+w_d) \bigr) \bigr|
&\le \ex_{u_1,\ldots,u_d\in U} \chi\bigl( T(u_1,\ldots,u_d)\bigr) \cr
&= \bias(T|_U) \cr
}$$
Since this is for all $w_1,\ldots, w_d$, averaging the left-hand side over all choices and
applying the triangle inequality gives us what we want.\slug

We are now able to show that for any tensor $T$ and independent set $A$ in $T$, we have
$\ark(T) \ge c|A|$ for $c$ depending on $d$ and $|\FF|$.

\proclaim Theorem \advthm. Let $T : (\FF^n)^d\to \FF$ be an order-$d$ tensor. Assume that
$A\subseteq [n]$ is an independent set in $T$. Then $\ark(T)\ge c|A|$ where
$$c = -\log_{|\FF|} \biggl( 1-\biggl( 1-{1\over y}\biggr)^{d-1}\biggr),$$
so that $c\ge 2^{-d}$ and $c\ge 1-\log(d-1)/\log |\FF|$.

\proof Let $S : (F^A)^d\to \FF$ be the restriction of $T$ to $\FF^A$.
By the previous lemma, $\ark(S)$ is a lower bound on $\ark(T)$, so we shall prove the theorem
by expressing $\ark(S)$ as a function of $|A|$. We have
$$\bias(A) = \ex_{x_1,\ldots,x_d\in \FF^A} \chi\Bigl( \sum_{i\in A} c_i \prod_{j\in [d]} x_{i,j}\Bigr),$$
where, since $A$ is an independent set, $c_i\ne 0$ for all $i\in A$. But we also know that
$$\bias(S) =
\pr_{x_2,\ldots,x_d\in \FF^A} \Bigl\{ \bigcap_{i\in A} \prod_{2\le j\le d} \{x_{i,j} = 0\}\Bigr\},$$
and this is easily seen to be
$$\bias(S) = \biggl( 1-\biggl( 1-{1\over |\FF|}\biggr)^{d-1} \biggr)^{|A|}.$$
Taking the logarithm of both sides, we see that $\ark(S) = c|A|$, where $c$ is as in the theorem
statement.

By convexity,
$$c \ge -\log_2 (1-2^{-(d-1)}) \ge 2^{-(d-1)}.$$
Also, as long as $|\FF|\ge d$, we have
$$c \ge -\log_{|\FF|}\bigl((d-1)/|\FF|\bigr) = 1-\log_{|\FF|} (d-1).\noskipslug$$
